### 2022-10-25 dissertation/chapters/01-introduction.tex
✨ (Chapter 3): PSO review done

A step further in the concept of hybridisation of learning paradigms is that of hybridisation of different \textit{heuristics} as they are applied to some optimisation problem~\cite{ref:burke:2013}. These methods are referred to as \acfp{HH} and focus on finding the best heuristic in \textit{heuristic-space} to solve a specific problem. One such form of \acs{HH} is a population-based approach that guides the search process by automatically selecting heuristics from a heuristic-pool to be applied to a collection of different candidate solutions in the solution-space. This collection of candidate solutions are referred to as a \textit{population} of \textit{entities}, where each \textit{entity} is a single candidate solution to the problem being optimised. Population-based \acp{HH} implement a strategy where multiple heuristics work together to solve a problem. This technique requires a mechanism that selects a specific heuristic to be applied to a specific candidate solution.

@@

One specific type of selection \acs{HH} is called a \index{multi-method populated-based meta-heuristic}\textit{multi-method population-based \index{meta-heuristic}meta-heuristic}~\cite{ref:vanderstockt:2018}. The term \index{multi-method}\textit{multi-method} refers to the incorporation of different low-level heuristics, with different search behaviours, into the heuristic-space. The term \textit{population-based} refers to the utilisation of a population of entities that represent candidate solutions. Finally, the term \index{meta-heuristic}\textit{meta-heuristics} refers to the \acs{HH} as a heuristic that does not have any domain knowledge and only makes use of information from the search process.

@@

\citeauthor{ref:grobler:2012}~\cite{ref:grobler:2012} mention that the focus of \acp{HH} is on automating the development of the \textit{learning mechanism} used to find the best heuristic to obtain an appropriate solution to an optimisation problem.~\citeauthor{ref:grobler:2012} explain that \acp{HH} employ a high-level heuristic that focuses on finding the best low-level heuristic in heuristic-space that could include \textit{single-method} and \textit{multi-method} optimisation algorithms (heuristics). \acp{HH} relieve the burden of having to select the best heuristic to use by trial and error. Furthermore, \acp{HH} can generally be applied to multiple problems given that the set of low-level heuristics include heuristics that have the potential to solve the problem at hand~\cite{ref:burke:2010}. Automation of the heuristic selection process and the general application to a wider range of problems are characteristics of \acp{HH} that can be beneficial when applied in the context of \acp{FFNN} training.

@@

      \item Conduct an empirical study and critically evaluate the performance of the developed \Acs{BHH} compared to individual heuristics in the heuristic-space as they are used to train \acp{FFNN} on a number of different problems.

>>>>

A step further in the concept of hybridisation of learning paradigms is that of hybridisation of different \textit{heuristics} as they are applied to some optimisation problem~\cite{ref:burke:2013}. These methods are referred to as \acfp{HH} and focus on finding the best heuristic in \textit{heuristic space} to solve a specific problem. One such form of \acs{HH} is a population-based approach that guides the search process by automatically selecting heuristics from a heuristic-pool to be applied to a collection of different candidate solutions in the solution-space. This collection of candidate solutions are referred to as a \textit{population} of \textit{entities}, where each \textit{entity} is a single candidate solution to the problem being optimised. Population-based \acp{HH} implement a strategy where multiple heuristics work together to solve a problem. This technique requires a mechanism that selects a specific heuristic to be applied to a specific candidate solution.

@@

One specific type of selection \acs{HH} is called a \index{multi-method populated-based meta-heuristic}\textit{multi-method population-based \index{meta-heuristic}meta-heuristic}~\cite{ref:vanderstockt:2018}. The term \index{multi-method}\textit{multi-method} refers to the incorporation of different low-level heuristics, with different search behaviours, into the heuristic space. The term \textit{population-based} refers to the utilisation of a population of entities that represent candidate solutions. Finally, the term \index{meta-heuristic}\textit{meta-heuristics} refers to the \acs{HH} as a heuristic that does not have any domain knowledge and only makes use of information from the search process.

@@

\citeauthor{ref:grobler:2012}~\cite{ref:grobler:2012} mention that the focus of \acp{HH} is on automating the development of the \textit{learning mechanism} used to find the best heuristic to obtain an appropriate solution to an optimisation problem.~\citeauthor{ref:grobler:2012} explain that \acp{HH} employ a high-level heuristic that focuses on finding the best low-level heuristic in heuristic space that could include \textit{single-method} and \textit{multi-method} optimisation algorithms (heuristics). \acp{HH} relieve the burden of having to select the best heuristic to use by trial and error. Furthermore, \acp{HH} can generally be applied to multiple problems given that the set of low-level heuristics include heuristics that have the potential to solve the problem at hand~\cite{ref:burke:2010}. Automation of the heuristic selection process and the general application to a wider range of problems are characteristics of \acp{HH} that can be beneficial when applied in the context of \acp{FFNN} training.

@@

      \item Conduct an empirical study and critically evaluate the performance of the developed \Acs{BHH} compared to individual heuristics in the heuristic space as they are used to train \acp{FFNN} on a number of different problems.

### 2022-10-25 dissertation/chapters/02-artificial-neural-networks.tex
✨ (Chapter 3): PSO review done

Weight initialisation is the process by which candidate solutions (represented by the weights of the \acs{AN}) to the problem are ``placed'' in the search-space. Weight initialisation influences the speed of convergence, the probability of convergence and the generalisation capabilities of \acp{ANN}~\cite{ref:fernandez:2001}. Finding the optimal initialisation values for weights is non-trivial and can be seen as another optimisation problem~\cite{ref:de:2016, ref:erdogmus:2003, ref:yam:2000}.

@@

Finding the optimal weights that produce the best performance on a given task is an optimisation problem. The optimisation algorithm used to find the optimal weights is referred to as a \index{heuristic}\textit{heuristic}. \index{heuristic}Heuristics search for possible solutions in the solution-space and make use of information from the search-space to guide to process. Heuristics are discussed in detail in Chapter~\ref{chap:heuristics}.

>>>>

Weight initialisation is the process by which candidate solutions (represented by the weights of the \acs{AN}) to the problem are ``placed'' in the search space. Weight initialisation influences the speed of convergence, the probability of convergence and the generalisation capabilities of \acp{ANN}~\cite{ref:fernandez:2001}. Finding the optimal initialisation values for weights is non-trivial and can be seen as another optimisation problem~\cite{ref:de:2016, ref:erdogmus:2003, ref:yam:2000}.

@@

Finding the optimal weights that produce the best performance on a given task is an optimisation problem. The optimisation algorithm used to find the optimal weights is referred to as a \index{heuristic}\textit{heuristic}. \index{heuristic}Heuristics search for possible solutions in the solution-space and make use of information from the search space to guide to process. Heuristics are discussed in detail in Chapter~\ref{chap:heuristics}.

### 2022-10-25 dissertation/chapters/03-heuristics.tex
✨ (Chapter 3): PSO review done

The biggest difference between \index{meta-heuristic}meta-heuristics and gradient-bases heuristics is that
\index{meta-heuristic}meta-heuristics make use of meta-information obtained as a result of evaluating the \acs{FFNN} during training and is not limited to information about the search-space~\cite{ref:blum:2003}.

@@

\Ac{PSO} is a stochastic population-based search algorithm based on the social behaviour of birds in a flock~\cite{ref:kennedy:1995}. By definition, the \acs{PSO} heuristic is nature-inspired.  \Acp{PSO} were first presented  by~\citeauthor{ref:kennedy:1995}\cite{ref:kennedy:1995}. Kennedy and Eberhart first applied \acp{PSO} to train of \acp{FFNN}~\cite{ref:eberhart:1995, ref:kennedy:1997}. The application of \acp{PSO} in the context of training \acs{FFNN} have been widely studied~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}. This section aims to provide the reader with the detail of \acs{PSO} implementation.

In general, this dissertation uses the term \textit{entity} for candidate solutions and a \textit{population} for a collection of entities.~\citeauthor{ref:engelbrecht:2007}\cite{ref:engelbrecht:2007} mentions that in \acp{PSO} individual candidate solutions are referred to as \textit{particles} and the population is referred to as a \textit{swarm}. These particles are ``flown'' through a hyperdimensional space. Changes in particle position is due to social-psychological tendencies of individuals to emulate the success of other individuals. The changes in the particle position are then influenced by the experience or knowledge of the particle's neighbours. The social behaviour of particles is modeled such that they stochastically return to previously successful regions in the search space.

@@

\noindent
\citeauthor{ref:shi:1998}~\cite{ref:shi:1998} proposed a modification of the original \acs{PSO} as was presented by~\citeauthor{ref:kennedy:1995}~\cite{ref:kennedy:1995}. Their implementation focuses on the \textit{gbest} \acs{PSO} with \textit{inertia} weights. This dissertation focuses on the \citeauthor{ref:shi:1998}~\cite{ref:shi:1998} implementation of \textit{global best} \acs{PSO}. The particles in this configuration has a number of properties associated with them~\cite{ref:vanwyk:2014}. These include:

\begin{itemize}
      \item \textbf{Position}: Refers to the candidate solution that is represented by the particle and defines the particle position within the optimisation problem's hyper-dimensional solution space. Let the current position of particle $i$ at timestep $t$ be denoted by $x_{i}(t)$ and let $N$ be the search space dimensionality.

@@

      \item \textbf{Solution Quality}: Refers to the evaluation of the particle's position with respect to the objective function. Let $f(x_{i}(t)$ denote the quality of the solution represented by the particle's position.

@@

\noindent
During initialisation, particles are randomly places within the search space by sampling from a uniform distribution such that $x_{i} \sim U(x_{min}, x_{max})$ and the velocity is set to 0. At timestep 0, the particle's initial position is set to be the particle's personal best solution such that $y_{i}(0) = x_{i}(0)$. The particle's update step is then broken into two parts. These include a velocity update step presented in Equation~\eqref{eq:heuristics:pso:velocity} followed by a position update step as presented in Equation~\eqref{eq:heuristics:pso:position} below.

\begin{equation}
      \label{eq:heuristics:pso:velocity}
      \begin{split}
            v_{ij}(t+1) = wv_{ij}(t) + c_{1}r_{1}(t)[y_{ij}(t) - x_{ij}(t)] + c_{2}r_{2}(t)[\hat{y}_{ij}(t) - x_{ij}(t)]
      \end{split}
\end{equation}

@@

      \item \textbf{Cognitive Component}: Denoted by the term $c_{1}r_{1}(t)[y_{ij} - x_{ij}(t)]$. This component represents the particle's personal  experience. It introduces an attractor to the particle's personal best position so far. The cognitive component is stochastically scaled with random numbers $r{1} \sim U(0,1)^N$ and the cognitive acceleration coefficient $c_{1}$ is used to control the influence of the cognitive attractor.

      \item \textbf{Social Component}: Denoted by the term $c_{2}r_{2}(t)[\hat{y}_{ij} - x_{ij}(t)]$. This component represents the particle's social  experience. It introduces an attractor to the swarm's best position so far. The social component is also stochastically scaled with random numbers $r{2} \sim U(0,1)^N$, while also introducing the social acceleration coefficient $c_{2}$ that is used to control the influence of the social attractor.
\end{itemize}

@@

\begin{algorithm}[H]
      \caption{The pseudo code algorithm for the gbest \acs{PSO} heuristic.}
      \label{algo:heuristics:pso:gbest}
      \begin{algorithmic}
            \State Create and initialise an $N_{x}$ dimensional swarm;
            \While{stopping condition not met}
            \For{each particle $i = 1, \dots, N_{s}$}
            \State // Set the personal best position;
            \If{$f(x_{i}) < f(y_{i})$ }
            \State $y_{i} = x_{i}$;
            \EndIf

@@

            \For{each particle $i = 1, \dots, N_{s}$}
            \State // Perform velocity update step;
            \State $v_{ij} = wv_{ij} + c_{1}r_{1}[y_{ij} - x_{ij}] + c_{2}r_{2}[\hat{y}_{ij} - x_{ij}]$
            \State // Perform position update step;
            \State $x_{ij} = x_{ij} + v_{ij}$
            \EndFor
            \EndWhile
      \end{algorithmic}
\end{algorithm}

>>>>

The biggest difference between \index{meta-heuristic}meta-heuristics and gradient-bases heuristics is that
\index{meta-heuristic}meta-heuristics make use of meta-information obtained as a result of evaluating the \acs{FFNN} during training and is not limited to information about the search space~\cite{ref:blum:2003}.

@@

\Acf{PSO} is a stochastic population-based search algorithm based on the social behaviour of birds in a flock~\cite{ref:kennedy:1995}. By definition, the \acs{PSO} heuristic is nature-inspired.  \Acp{PSO} were first presented  by~\citeauthor{ref:kennedy:1995}\cite{ref:kennedy:1995}. Kennedy and Eberhart first applied \acp{PSO} to train of \acp{FFNN}~\cite{ref:eberhart:1995, ref:kennedy:1997}. The application of \acp{PSO} in the context of training \acs{FFNN} have been widely studied~\cite{ref:rakitianskaia:2012, ref:vanwyk:2014}. This section aims to provide the reader with the detail of \acs{PSO} implementation.

This dissertation uses the term \textit{entity} for candidate solutions and a \textit{population} for a collection of entities, in the general context of population-based \index{meta-heuristic}meta-heuristics.~\citeauthor{ref:engelbrecht:2007}~\cite{ref:engelbrecht:2007} mentions that in \acp{PSO} individual candidate solutions are referred to as \textit{particles} and the population is referred to as a \textit{swarm}. These particles are ``flown'' through a hyperdimensional search space. Changes in particle position is due to social-psychological tendencies of individuals to emulate the success of other individuals. The changes in the particle position are then influenced by the experience or knowledge of the particle's neighbours. The social behaviour of particles is modeled such that they stochastically return to previously successful regions in the search space.

@@

\noindent
\citeauthor{ref:shi:1998}~\cite{ref:shi:1998} proposed a modification of the original \acs{PSO} as was presented by~\citeauthor{ref:kennedy:1995}~\cite{ref:kennedy:1995}. Their implementation focuses on the \textit{gbest} \acs{PSO} with \textit{inertia} weights. This dissertation focuses on the \citeauthor{ref:shi:1998}~\cite{ref:shi:1998} implementation of \textit{global best} \acs{PSO} and particles in this specific implementation has a number of properties associated with them~\cite{ref:vanwyk:2014}. These include:

\begin{itemize}
      \item \textbf{Position}: Refers to the candidate solution that is represented by the particle and defines the particle position within the optimisation problem's hyper-dimensional solution space. Let the current position of particle $i$ at timestep $t$ be denoted by $x_{i}(t)$ and let $J$ be the search space dimensionality.

@@

      \item \textbf{Solution Quality}: Refers to the evaluation of the particle's position with respect to the objective function. Let $f(x_{i}(t))$ denote the quality of the solution represented by the particle's position.

@@

\noindent
During initialisation, particles are randomly places within the search space by sampling from a uniform distribution such that $x_{i} \sim U(x_{min}, x_{max})$ and the velocity is set to 0. At timestep 0, the particle's initial position is set to be the particle's personal best solution such that $y_{i}(0) = x_{i}(0)$. The particle's update step is then broken into two parts, including a velocity update step presented in Equation~\eqref{eq:heuristics:pso:velocity} followed by a position update step as presented in Equation~\eqref{eq:heuristics:pso:position} below.

\begin{equation}
      \label{eq:heuristics:pso:velocity}
      \begin{split}
            v_{ij}(t+1) = wv_{ij}(t) + c_{1}r_{1_{j}}(t)[y_{ij}(t) - x_{ij}(t)] + c_{2}r_{2_{j}}(t)[\hat{y}_{ij}(t) - x_{ij}(t)]
      \end{split}
\end{equation}

@@

      \item \textbf{Cognitive Component}: Denoted by the term $c_{1}r_{1_{j}}(t)[y_{ij} - x_{ij}(t)]$. This component represents the particle's personal  experience. It introduces an attractor to the particle's personal best position so far. The cognitive component is stochastically scaled with random numbers $r_{1} \sim U(0,1)^J$ and the cognitive acceleration coefficient $c_{1}$ is used to control the influence of the cognitive attractor.

      \item \textbf{Social Component}: Denoted by the term $c_{2}r_{2+{j}}(t)[\hat{y}_{ij} - x_{ij}(t)]$. This component represents the particle's social  experience. It introduces an attractor to the swarm's best position so far. The social component is also stochastically scaled with random numbers $r_{2} \sim U(0,1)^J$, while also introducing the social acceleration coefficient $c_{2}$ that is used to control the influence of the social attractor.
\end{itemize}

@@

\begin{algorithm}[H]
      \caption{The pseudo code algorithm for the gbest \acs{PSO} heuristic.}
      \label{algo:heuristics:pso:gbest}
      \begin{algorithmic}
            \State Create swarm of $N$ entities, each with $J$ dimensions;
            \While{stopping condition are not met}
            \For{each particle $i = 1, \dots, N$}
            \State // Set the personal best position;
            \If{$f(x_{i}) < f(y_{i})$ }
            \State $y_{i} = x_{i}$;
            \EndIf

@@

            \For{each particle $i = 1, \dots, N$}
            \For{each dimension $j = 1, \dots, J$}
            \State // Perform velocity update step;
            \State $v_{ij} = wv_{ij} + c_{1}r_{1_{j}}[y_{ij} - x_{ij}] + c_{2}r_{2_{j}}[\hat{y}_{ij} - x_{ij}]$
            \State // Perform position update step;
            \State $x_{ij} = x_{ij} + v_{ij}$
            \EndFor
            \EndFor
            \EndWhile
      \end{algorithmic}
\end{algorithm}

### 2022-10-25 dissertation/chapters/04-hyper-heuristics.tex
✨ (Chapter 3): PSO review done

Importantly, \ac{HH} are concerned with finding the best heuristic in heuristic-space, while the underlying low-level heuristics find solutions in the feasible search space. This level of abstraction is illustrated in Figure \ref{fig:heuristics:hh:hh_algorithm} below.

>>>>

Importantly, \ac{HH} are concerned with finding the best heuristic in heuristic space, while the underlying low-level heuristics find solutions in the feasible search space. This level of abstraction is illustrated in Figure \ref{fig:heuristics:hh:hh_algorithm} below.

### 2022-10-25 dissertation/chapters/06-bayesian-hyper-heuristic.tex
✨ (Chapter 3): PSO review done

      \item \textbf{Online Learning:} Refers to the source of the feedback used during training. The \ac{BHH} optimises, at a higher level, the selection and application of the \index{heuristic-space}\textit{heuristic-space} while low-level heuristics optimise entities' solution to the underlying model, in this case a \ac{FFNN}. This happens during training time, yielding the classification of utilising \textit{online} learning.
\end{itemize}

>>>>

      \item \textbf{Online Learning:} Refers to the source of the feedback used during training. The \ac{BHH} optimises, at a higher level, the selection and application of the \index{heuristic space}\textit{heuristic space} while low-level heuristics optimise entities' solution to the underlying model, in this case a \ac{FFNN}. This happens during training time, yielding the classification of utilising \textit{online} learning.
\end{itemize}

### 2022-10-25 dissertation/chapters/08-results.tex
🚧 WIP - Save point (compiles)
✨ (Chapter 3): PSO review done

One can then conclude that most of the overfitting occurs as a result of the inclusion of the meta-heuristics. To some extent this does make sense, since the proxied update step operations for these heuristics do not map as well between meta-heuristics as they do with gradient-based heuristics and thus an update step that occurs when switching from a \acs{PSO} to a \acs{GA} for example might maintain a state of momentum in a certain direction and then as a result of cross-over end up in a totally opposite direction, resulting in the entity vastly overshooting any optima. Van Wyk \cite{ref:vanwyk:2014} did a detailed analysis of overfitting that occurs when training \acp{FFNN} using \acp{PSO}. Although the same velocity clamping techniques were used as suggested in Van Wyk's work, they are not guaranteed to avoid overfitting. Another technique would be to nullify and reset velocites and momentum, but this could then be detrimental to the other heuristics. Since the \Ac{BHH} includes a \acs{PSO} in the heuristic pool, one could expect similar results.

@@

Credit should however be given to the \Ac{BHH} for its ability to learn positively in such a short timespan. It has been indicated that the update step for the \Ac{BHH} happens at each mini-batch iteration. If a large dataset is presented, that means that the majority of training takes place in a small number of epochs. This leaves a very small window for the \Ac{BHH} to actually learn relevant and useful information from the \index{heuristic}heuristic-space.

>>>>

One can then conclude that most of the overfitting occurs as a result of the inclusion of the meta-heuristics. To some extent this does make sense, since the proxied update step operations for these heuristics do not map as well between meta-heuristics as they do with gradient-based heuristics and thus an update step that occurs when switching from a \acs{PSO} to a \acs{GA} for example might maintain a state of momentum in a certain direction and then as a result of cross-over end up in a totally opposite direction, resulting in the entity vastly overshooting any optima. Van Wyk~\cite{ref:vanwyk:2014} did a detailed analysis of overfitting that occurs when training \acp{FFNN} using \acp{PSO}. Although the same velocity clamping techniques were used as suggested in Van Wyk's work, they are not guaranteed to avoid overfitting. Another technique would be to nullify and reset velocites and momentum, but this could then be detrimental to the other heuristics. Since the \Ac{BHH} includes a \acs{PSO} in the heuristic pool, one could expect similar results.

@@

Credit should however be given to the \Ac{BHH} for its ability to learn positively in such a short timespan. It has been indicated that the update step for the \Ac{BHH} happens at each mini-batch iteration. If a large dataset is presented, that means that the majority of training takes place in a small number of epochs. This leaves a very small window for the \Ac{BHH} to actually learn relevant and useful information from the \index{heuristic}heuristic space.
